Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-10-15 docs/papers/shallow_performance/shallow_perf.tex
parco paper: changes following comments by Eike

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal\, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache-based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see performance
improvement) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system. Further, the original code performed poorly using the
IBM compiler on Power8 and needed to be modified to obtain performant
code.
%%
Therefore, the \psykal\ approach can be used with negligible
performance loss and sometimes small performance gains compared to the
original optimised code. We also find that there is no single best
hand-optimised implementation of the code for all of the compilers
tested.

@@

The bottom layer, in terms of calling hierarchy, is the Kernel
layer. The Kernel layer implements the science that the Algorithm
layer calls, as a set of subroutines. These kernels operate on fields
that are local to the process doing the computation. (Depending on the
type of kernel, these may be a set of elements, a single column of
elements, or a set of columns.). Again the scientist is responsible
for writing this layer and there is no parallelism specified here,
but, depending on the complexity of the Kernels, there may be input
from an HPC expert and/or some coding rules to help ensure the kernels
compile into efficient code.

@@

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer
can therefore be tailored for a particular hardware (such as
multi-core, many-core, GPGPUs, or some combination thereof) and
software (such as compiler, operating system, MPI implementation)
configuration with no change to the Algorithm or Kernel layer
code. This approach therefore offers the potential for portable
performance.

@@

Excluding the application of boundary conditions, the code has three
main sections: computation of mass fluxes, vorticity and sea-surface
height; computation of the pressure and velocity components; and
finally, smoothing of the fields in time and copying of the field
values at the next time-step to be those at the current
time-step. Profiling of the original code showed that these 
sections each account for approximately one third of the run-time.

@@

As indicated above, transformations T10 and T11 were
suggested by the IBM compiler team. The second of these effectively
undoes transformation four by fissioning the first loop nest back
into three separate loop nests.

We shall see that these transformations do not always result in
improved performance. Whether or not they do so depends both on the
compiler and the problem size. We also emphasise that the aim of these
optimisations is typically to recover, as far as is possible, the
structure of the original version of the code. It may well be that
transforming the code into some other structure would result in better
performance on a particular architecture. However, exploring this
optimisation space is beyond the scope of the present work.

@@

Strangely, the third bar (Gnu compiler) in
Figure~\ref{FIG_orig_perf_summary} shows the same drop in performance
in moving from the $256^2$ case to the $512^2$ case, despite the
larger LLC of the Ivy Bridge system. In order to investigate this
further, we profiled the application to obtain hardware performance
counter data for the Gnu- and Intel-compiled executables. This
revealed that the Gnu-compiled executable made more references to
memory and had a higher LLC miss-rate compared with the Intel binary.
We conclude that the Gnu compiler is not utilising registers as
efficiently as the Intel compiler and that this results in greater
memory traffic.

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}). We acknowledge use of the Hartree
Centre IBM Power 8 system in this work. The STFC Hartree Centre is a
research collaboratory in association with IBM providing High
Performance Computing platforms funded by the UK's investment in
e-Infrastructure. The Centre aims to develop and demonstrate next
generation software, optimised to take advantage of the move towards
exa-scale computing. This work was funded by the NERC `GOcean' project,
grant number NE/L012111/1.

>>>>

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal\, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache-based
architectures is taken as the starting point. A vanilla
\psykal\ version is written and the performance of the two
compared. The optimisations that were applied to the original
benchmark (loop fusion {\it etc.}) are then manually applied to the
\psykal\ version as a set of code modifications to the optimisation
layer. Performance results are presented for the Cray, Intel and GNU
compilers on Intel Ivybridge and Haswell processors and for the IBM
compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see performance
improvement) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system. Further, the original code performed poorly using the
IBM compiler on Power8 and needed to be modified to obtain performant
code.
%%
Therefore, the \psykal\ approach can be used with negligible
performance loss and sometimes small performance gains compared to the
original optimised code. We also find that there is no single best
hand-optimised implementation of the code for all of the compilers
tested.

@@

The bottom layer, in terms of calling hierarchy, is the Kernel
layer. The Kernel layer implements the science that the Algorithm
layer calls, as a set of subroutines. These kernels operate on fields
that are local to the process doing the computation. (Depending on the
type of kernel, these may be a set of elements, a single column of
elements, or a set of columns.). Again the scientist is responsible
for writing this layer and there is no parallelism specified here,
but, depending on the complexity of the Kernels, there may be input
from an HPC expert and/or some coding rules to help ensure the kernels
compile into efficient code.  In an alternative approach, kernels are
generated from a high-level specification~\cite{firedrake,fenics},
potentially allowing them to be optimised automatically~\cite{coffee}.

@@

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer can
therefore be tailored for a particular hardware (such as multi-core,
many-core, GPGPUs, or some combination thereof) and software (such as
compiler, operating system, MPI implementation) configuration with no
change to the Algorithm or Kernel layer code. This approach therefore
offers the potential for portable performance. In this work we apply
optimisations to the PSy layer manually. The development of a tool to
automate this process will be the subject of a future paper.

@@

Excluding the application of boundary conditions, the code has three
main sections: computation of mass fluxes, vorticity and sea-surface
height; computation of the pressure and velocity components; and
finally, smoothing of the fields in time and copying of the field
values at the next time-step to be those at the current
time-step. Profiling the original code showed that these 
sections each account for around one third of the run-time.

@@

As indicated above, transformations T10 and T11 were suggested by the
IBM compiler team. The second of these effectively undoes
transformation four by fissioning the first loop nest back into three
separate loop nests. Note that this sequence of transformations is
specific to Shallow: we are currently working upon a code-generation
system that will allow a user to choose from a menu of transformations
when optimising a PSy layer and this will be the subject of a future
paper.

We shall see that the transformations T1--T11 do not always result in
improved performance. Whether or not they do so depends both on the
compiler and the problem size. We also emphasise that the aim of these
optimisations is typically to recover, as far as is possible, the
structure of the original version of the code.  Therefore this
sequence of transformations is specific to Shallow. It may well be
that transforming the code into some other structure would result in
better performance on a particular architecture. However, exploring
this optimisation space is beyond the scope of the present work.  We
are currently developing a code-generation system that will allow a
user to choose from a menu of transformations when optimising a PSy
layer and this will be the subject of a future paper.

@@

Strangely, the third bar (Gnu compiler) in
Figure~\ref{FIG_orig_perf_summary} shows the same drop in performance
in moving from the $256^2$ case to the $512^2$ case, despite the
larger LLC of the Ivy Bridge system. In order to investigate this, we
profiled the application to obtain hardware performance counter data
for the Gnu- and Intel-compiled executables. This revealed that the
Gnu-compiled executable made more references to memory and had a
higher LLC miss-rate compared with the Intel binary.  We conclude that
the Gnu compiler is not utilising registers as efficiently as the
Intel compiler and that this results in greater memory traffic.

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}) and the Hartree Centre IBM Power 8
system. The STFC Hartree Centre is a research collaboratory in
association with IBM providing HPC platforms funded by the UK's
investment in e-Infrastructure. The Centre aims to develop and
demonstrate next generation software, optimised for
the move towards exa-scale computing. This work was funded by the NERC
`GOcean' project, grant number NE/L012111/1.
